Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-08 text/manuscript-full-KS.qmd
front end revise

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Scholarship from the economics of education often investigate financial aid leveraging, which seeks to convert admits to enrolled students [e.g., @RN3762; @RN3133]. <!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

@@

By focusing on the behavior of colleges, scholarship on EM implicitly states that recruiting is something done by individual colleges. The broader EM industry includes relevant stakeholders in the organizational field [@RN527], including professional associations (e.g., National Association for College Admission Counseling) and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges. We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Although sociologists have hinted at the ways enrollment management contributes to inequality in college access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  --> <!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory, enrollment management processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. Unfortunately, few empirical studies make EM vendors or products the object of empirical analysis. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

@@

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The Student Search Service -- created in 1972 and copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification. Student Search Service includes or excludes prospects from a student list purchase based on value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on equality of opportunity.

@@

@RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402). Reacivity, discipline, and authority are three interrelated themes that describe the effects of quantification. *Reactivity* is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification *disciplines* actors to react in particular ways. Quantification changes decision-making power *authority*, often weakening the discretion of local decision-makers. For instance, scholarship about U.S. News & World Report (USNWR) law school rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll [@RN5003]. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions personnel to make independent decisions about class size and which applicants to admit [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools.<!-- THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```
**Correlation**. The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4772; @RN4849]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and Geomarkets. Correlation measures the extent to which two or more variables move together. Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that

@@

Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

@@

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homopphily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. YY]. We observe similar phenomena in our case. Based on a snapshot of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

@@

We make targeted contributions to scholarship on quantification in the sociology of education. First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375]. As @RN3519 demonstrates in the chapter *Numbers*, EM is fundamentally concerned with quantification. However, sociological analyses of enrollment management focus on the behavior of colleges and their agents, yielding the implicit belief that inequalities created by EM are a function of individual colleges reacting to rankings and other macro structures in their environment. To our knowledge, the sociology of education has not considered producer-facing products that quantify students on behalf of producers. Market research and investigative journalism indicates that the vast majority of colleges purchase third-party EM products to identify prospects and to decide which schools and communities to target [@RN5027; ]. Drawing from scholarship from critical data studies [e.g., @RN4775; @RN4975; @RN4849], recruiting products grounded in the logic of predictive analytics take a snapshot of student demand -- without considering historical structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches verticially categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Third-party EM products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-service platforms, but have remained invisible to sociologists. We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables, @RN5005 we cannot show the effect of quantification on social reproduction. However, by showing which students are included/excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in educational opportunity.

@@

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting (p.9)\ldots We have begun with this celebration of sotrytelling\ldots \[because\] we believe that the intuitios of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research this is based opn listening carefully to what admissions officers have to say (pp. 9-10).

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one anothyer.[^7] Based on these analyses, @RN4982 \[p. 46\] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

@@

**Enrollment Planning Service**. In 1984, College Board created the Enrollment Planning Service (EPS), based on the Market Segment Model [@RN4987; @RN4980]. EPS was an early software-as-service platform that recreated the analyses of @RN4982. For each Geomarket, colleges could obtain Market Segment Report for each local market and the Institutional Profile -- their own and that of competitors. Based on background conversations with enrollment management professionals, EPS software also provided information about the score-sending behavior of individual high schools within each Geomarket. Therefore, colleges used EPS software to decide which Geomarkets to recruit from and which high schools to visit within targeted Geomarkets. Typical @RN5019 marketing material describes EPS as,

@@

EPS may also weaken the authority of local decision-makers. @RN4982 sought to develop a concrete, data-driven framework -- The Market Segment Model -- that replicates the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an enrollment management consultant -- to plan recruiting efforts centrally. In background conversations, enrollment management professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools.

@@

The discussion of homophily by @RN4975 describe the logic of The Market Segment model but also an underlying institutional logic of the enrollment management industry. The logic of homophily argues that there is a natural cultural match between students, that colleges should find students who look like students who enrolled at their institution in the past, and that colleges should recruit from the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The also find great overlap in the schools visited by Villanova and Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

>>>>

Scholarship at the nexus of EM and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Scholarship from the economics of education often investigate financial aid leveraging, which seeks to convert admits to enrolled students [e.g., @RN3762; @RN3133]. <!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

@@

By focusing on the behavior of colleges, scholarship on EM implicitly assumes that recruiting is something done by individual colleges. The broader EM industry includes relevant stakeholders in the organizational field [@RN527], including professional associations (e.g., National Association for College Admission Counseling) and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges. We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Although sociologists have hinted at the ways EM contributes to inequality in college access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  --> <!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. Unfortunately, few empirical studies make EM vendors or products the object of empirical analysis. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

@@

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The Student Search Service -- created in 1972 and copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from a student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

Quantification can be defined as "the production and communication of numbers" [@RN5007, p. 402]. While a growing interdisciplinary area of scholarship, sociological studies on quantification have predominantly explored how numbers are produced, used, and interpreted across institutional contexts such as administration, democratic rule, economics, and personal life [@RN5008]. Quantification has also been at the core of recent critical data scholarship exploring how algorithms and big data computation structure social opportunities [e.g., @RN4772; @RN4849]. Our study is grounded in the systematic investigation of "interactions between different quantification regimes" across sociology and critical data studies, as called for by @RN5008 (p. 239), to explore the role of third-party servicers and products in the sorting of students into colleges. Specifically, we draw on quantification scholarship investigating how correlation and homophily [@RN4975] encode social reproduction into our growingly dominant "algorithmic society" \[[CITE](https://medium.com/chatbots-bible-and-more/theres-an-algorithm-for-that-or-there-soon-will-be-d54406dc9bc1)\] and explore how such mechanisms are applied to the construction of Geomarkets in the Market Segment Model [@RN4982].

<!--Reactivity, discipline, and authority are three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification disciplines actors to react in particular ways. Quantification changes decision-making power authority, often weakening the discretion of local decision-makers. For instance, scholarship about U.S. News & World Report (USNWR) law school rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll [@RN5003]. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- used rankings to make decisions that had large consequences for schools \ldots [then] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings disciplined the behavior of law schools. For example, characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the authority of admissions personnel to make independent decisions about class size and which applicants to admit [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

<!--The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools. THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```

<!--# REMOVED FROM PARAGRAPH BELOW: The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4772; @RN4849]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and Geomarkets.  -->

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that

@@

Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with an variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

@@

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homophily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. YY]. We observe similar phenomena in our case. Based on a snapshot of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

@@

We make targeted contributions to scholarship on quantification in the sociology of education. As @RN3519 demonstrates in the chapter *Numbers*, EM is fundamentally concerned with quantification. However, sociological analyses of EM focus on the behavior of colleges and their agents, yielding the implicit belief that inequalities created by EM are a function of individual colleges reacting to macro structures in their environment such as rankings and financial pressures. To our knowledge, the sociology of education has not considered producer-facing products that quantify students on behalf of producers. Market research and investigative journalism indicates that the vast majority of colleges purchase third-party EM products to identify prospects and to decide which schools and communities to target [@RN5027]. Drawing from critical data studies [e.g., @RN4775; @RN4975; @RN4849], recruiting products grounded in the logic of predictive analytics take a snapshot of student demand -- without considering historical structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches vertically categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Third-party EM products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-service platforms, but have remained invisible to sociologists. We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.

<!--# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. By showing which students are included versys excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

<!--# removed from paragraph above: Unlike the analysis of UK school league tables, @RN5005 we cannot show the effect of quantification on social reproduction. However, b -->

@@

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting (p.9)\ldots We have begun with this celebration of storytelling\ldots \[because\] we believe that the intuitions of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research is based upon listening carefully to what admissions officers have to say (pp. 9-10).

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one another.[^7] Based on these analyses, @RN4982 \[p. 46\] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

@@

**Enrollment Planning Service**. In 1984, College Board created the Enrollment Planning Service (EPS), based on the Market Segment Model [@RN4987; @RN4980]. EPS was an early software-as-service platform that recreated the analyses of @RN4982. For each Geomarket, colleges could obtain Market Segment Report for each local market and the Institutional Profile -- their own and that of competitors. Based on background conversations with EM professionals, EPS software also provided information about the score-sending behavior of individual high schools within each Geomarket. Therefore, colleges used EPS software to decide which Geomarkets to recruit from and which high schools to visit within targeted Geomarkets. Typical @RN5019 marketing material describes EPS as,

@@

EPS may also weaken the authority of local decision-makers. @RN4982 sought to develop a concrete, data-driven framework -- The Market Segment Model -- that replicates the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools.

@@

The discussion of homophily by @RN4975 describe the logic of The Market Segment model but also an underlying institutional logic of the EM industry. The logic of homophily argues that there is a natural cultural match between students, that colleges should find students who look like students who enrolled at their institution in the past, and that colleges should recruit from the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The also find great overlap in the schools visited by Villanova and Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.
